A replication of a simulation study on measuring and correcting bias of under-5 mortality estimation in populations affected by HIV/AIDS

### 2020-02-24 analysis/paper.Rmd
update

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '../') 
```
# Methods

```{r, load-packages, include=FALSE}
if(!require(tidyverse)){
  install.packages('http://cran.r-project.org/src/contrib/Archive/tidyverse/tidyverse_1.2.1.tar.gz',repos=NULL, type="source")
}
packageVersion("tidyverse")
require(tidyverse)
# ── Attaching packages ───────────────────────────────────────────────────────────────────────────── tidyverse 1.2.1 ──
# ✔ ggplot2 3.1.0       ✔ purrr   0.3.0  
# ✔ tibble  2.0.1       ✔ dplyr   0.8.0.1
# ✔ tidyr   0.8.2       ✔ stringr 1.4.0  
# ✔ readr   1.1.1       ✔ forcats 0.3.0  
# ── Conflicts ──────────────────────────────────────────────────────────────────────────────── tidyverse_conflicts() ──
# ✖ dplyr::filter() masks stats::filter()
# ✖ dplyr::lag()    masks stats::lag()

if(!require(foreach)){
  install.packages('https://cran.r-project.org/src/contrib/Archive/foreach/foreach_1.4.4.tar.gz',repos=NULL, type="source")
}
packageVersion("foreach")
# 1.4.4
require(foreach)

if(!require(iterators)){
  install.packages('https://cran.r-project.org/src/contrib/Archive/iterators/iterators_1.0.10.tar.gz',repos=NULL, type="source")
}
packageVersion("iterators")
# 1.0.10
require(iterators)

if(!require(bcaboot)){
  install.packages("bcaboot",dependencies = TRUE,repos='http://cran.us.r-project.org')
}
require(bcaboot)
packageVersion("bcaboot")
# 0.2.1

if(!require(data.table)){
  install.packages("data.table",dependencies = TRUE,repos='http://cran.us.r-project.org')
}
require(data.table)
packageVersion("data.table")
# 1.11.8

if(!require(pls)){
  install.packages("pls",dependencies = TRUE,repos='http://cran.us.r-project.org')
}
require(pls)
packageVersion("pls")
# 2.7.0

if(!require(glmnet)){
  install.packages('https://cran.r-project.org/src/contrib/Archive/glmnet/glmnet_2.0-16.tar.gz',repos=NULL, type="source")
}
require(glmnet)
packageVersion("glmnet")
# 2.0.16

if(!require(officer)){
  install.packages("officer",dependencies = TRUE,repos='http://cran.us.r-project.org')
}
require(officer)
packageVersion("officer")
# 0.3.2

if(!require(flextable)){
  install.packages("flextable",dependencies = TRUE,repos='http://cran.us.r-project.org')
}
require(flextable)
packageVersion("flextable")
# 0.5.1

if(!require(magrittr)){
  install.packages("magrittr",dependencies = TRUE,repos='http://cran.us.r-project.org')
}
require(magrittr)
packageVersion("magrittr")
# 1.5

@@

For the calculation of bias, we used the formula provided in the original paper. 
$$ Relative bias = {\frac{IE_{survivors} - IE_{survivors&HIV deaths}}{IE_{survivors&HIV deaths}} $$        
$$ Absolute bias = IE_{survivors} - IE_{survivors&HIV deaths}$$    
According to [@quattrochi2019], ${IE_{survivors}$ was the estimates of U5M based on living women in 2010, and IE_{survivors&HIV deaths} was the estimates of U5M based on living women in 2010 and women died from HIV/AIDS before 2010 given the condition that they would have been 15-49 in 2010 if they had survived.

## Simulation Construction 
```{r, include = FALSE, eval=FALSE}
##### DEFINE FOLDERS #######

if(!dir.exists("results/figdata")){
  dir.create("results/figdata",recursive = TRUE)
}

if(!dir.exists("results/regdata")){
  dir.create("results/regdata",recursive = TRUE)
}

# Simulation results will be stored here

if(!dir.exists("results/models")){
  dir.create("results/models",recursive = TRUE)
}

##### SIMULATION #####

# Read data

hivhogan = read.csv("data/inc_curves.csv",head=TRUE)
mort_series = read.csv("data/IHME_female_mortSMALL.csv",head=TRUE)
adultmort = read.csv("data/MLTfemSMALL.csv",head=TRUE)
worldfert= read.csv("data/world_fert.csv",head=TRUE)
tfr_series= read.csv("data/tfr_gapminder_long.csv",head=TRUE)
art_series= read.csv("data/sampleART.csv",head=TRUE)
u5m_edit= read.csv("data/u5m_edit.csv",head=TRUE)
matmort = read.csv("data/matmort.csv",head=TRUE)

# Add HIV-free populations to hivhogan
hivhogan$Country <- factor(hivhogan$Country,levels=c(levels(hivhogan$Country),"zero"))
hivhogan[63,1] = "zero"
hivhogan[63,c(2:47)]=0

# Create parameters sets

fertcountry <- c("Botswana","Uganda")
cm_cntry <-  c("Mali","Morocco")
am_cntry <-  c("Madagascar","Sudan")
sexactive15 <-  c(30,70)
mmr0 <- c(0.0012,0.012)
mmr_dec <- c(0,0.073) # annual percent decline in MMR
curve <- c("BotswanaUrban","LesothoRural","MalawiRural","UgandaRural","CamerounRural","zero","BotUrb2x")
bfeed <- c(6,18)
art_col <- c("zero","Botswana","Cameroon","Malawi","Bot_dub")
growth <-  c(0.03)
yrend <- 2010
# Starting year
yrstart <- 1906

# CD4 threshold

threshold <- 200

# expand.grid() will create dataset with unique combinations in each row

inputs <- expand.grid(fertcountry, cm_cntry, am_cntry, sexactive15,mmr0,mmr_dec,curve,bfeed,art_col,growth,yrend,yrstart,threshold)
names(inputs) = c("fertcountry", "cm_cntry","am_cntry","sexactive15","mmr0","mmr_dec","curve","bfeed","art_col","growth","yrend","yrstart","threshold")

# Save the inputs set for later

save(inputs,file = "results/models/inputs.RData")

file_number_format <- paste0("%0",nchar(as.character(nrow(inputs))),"d")

# Set size of initial population

initial_populations<- c(7500,15000,22500)
years <- c(1906:2010)
ages <- c(0:120)

# Run the simulations for each initial population

for(ip in initial_populations){

  # Parameter sets already processed
  #ip = 150
  psets_already <- as.integer(gsub("models|\\.Rdata","",list.files(paste0("./results/models/p",ip),pattern = "models\\d+",full.names = FALSE)))

  psets_to_process <- setdiff(1:nrow(inputs),psets_already)

  # Run simulation for each set of parameters

  foreach(pset=psets_to_process,.packages = c("dplyr")) %dopar%{

    inp <- inputs[pset,]

    source("R/simulation_functions.R")

    source("R/indirect_estimates_functions.R")

    # Run simmulation
    set.seed(1)
    results <- bigsim(inp,initialpop=ip,years,ages,hivhogan,mort_series,adultmort,worldfert,tfr_series,art_series,u5m_edit,matmort)

    # Save simulation results
    save(results,file=file.path(paste0("./results/models/p",ip),paste("models",sprintf(file_number_format,pset),".Rdata",sep="")))
  }

# Collect hiv2000 differences among initial populations

# For each input row

hiv2000 <- foreach(pset=1:1000,.packages = c("dplyr")) %dopar%{

  # And each initial population
  paste0("p",initial_populations) %>% lapply(function(popini){

    # Load the results file for that initial population and input set

    load(file.path("results/models/",popini,paste("models",sprintf(file_number_format,pset),".Rdata",sep="")))
    print(paste(initial_populations,pset))
    # Extract the value of hiv2000
    r <- results[[1]]$hiv2000

    names(r) <- popini
    r
  }) %>% unlist() %>% t %>% data.frame(input_set=pset,.) # Put everything in one data.frame

} %>% bind_rows

# Compute difference

hiv2000 <- hiv2000 %>% mutate(p7500_minus_p15000=p7500-p15000,p15000_minus_p22500=p15000-p22500)

save(hiv2000,file = "results/hiv2000.RData")

##### INDIRECT ESTIMATES FOR INITAL POPULATION 22500 #####

# See if we have something not processed

psets_already <- intersect(as.integer(gsub("figdata.|\\.Rdata","",list.files("./results/figdata/p22500",pattern = "figdata.\\d+",full.names = FALSE))),
                           as.integer(gsub("regdata.|\\.Rdata","",list.files("./results/regdata/p22500",pattern = "regdata.\\d+",full.names = FALSE))))

psets_to_process <- setdiff(1:nrow(inputs),psets_already)

# For each input set not processed

foreach(pset=1:1000,.packages = c("dplyr")) %dopar%{

  source("R/indirect_estimates_functions_revised.R")

  load(file.path("results/models/p22500/",paste("models",sprintf(file_number_format,pset),".Rdata",sep="")))
  print(pset)

  w <- results[[7]]

  # Are babies of HIV moms getting HIV?
  # What percent of HIV moms babies have HIV?
  # How many born with HIV? 
  # df = as.data.frame(w)
  # y <- subset(df, hiv_date==2009)
  # plot(df$dob, df$hiv_date)  
  ##################################################

  # format data
  momkidsclean <- as.data.frame(w)

  # Using all surviving women
  isf <-ind_est(momkidsclean)

  # Using all women
  isf_all <-ind_est_all(momkidsclean)

  #Using surviving women and women who died from HIV
  isf_hiv <- ind_est_hiv(momkidsclean)

  ie.three <- cbind(isf,isf_all,isf_hiv,pset)

  ###################################################################
  # Save results for regressions and for figures
  ###################################################################

  inp.plus <- as.data.frame(results[1])    
  regdata <- cbind(ie.three,inp.plus[1,], row.names = NULL) # Data for regressions
  figdata <- results[c(2:6,8:9)] # Data for figures

  save(regdata,file=file.path("results/regdata/p22500/",paste("regdata.",sprintf(file_number_format,pset),".Rdata",sep="")))
  save(figdata,file=file.path("results/figdata/p22500/",paste("figdata.",sprintf(file_number_format,pset),".Rdata",sep="")))
  print('Saved')
  rm(results,regdata,figdata,momkidsclean,isf,isf_all,isf_hiv,ie.three)
}

##### COLLECT DATA FROM ALL SIMULATIONS FOR INITIAL POPULATION 22500 #####

# Merge regdata

psets_to_process <- as.integer(gsub("regdata.|\\.Rdata","",list.files("./results/regdata/p22500",pattern = "regdata.\\d+",full.names = FALSE))) %>% sort

nbd2k <- psets_to_process %>% lapply(function(pset){

  filename <- file.path("results/regdata/p22500",paste("regdata.",sprintf(file_number_format,pset),".Rdata",sep=""))
  tryCatch({
    load(filename)

    regdata %>% mutate(i=pset) %>% select(i,everything())

  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}) %>% bind_rows() %>% as.data.frame()

save(nbd2k,file="results/regdata/p22500/regdata_all.Rdata")

# Merge figdata

h <- list()
tf<- list()
ar<- list()
hd<- list()
u5m <- list()
inps <- list()
arp <- list()
hivinc <- list()

psets_to_process <- as.integer(gsub("figdata.|\\.Rdata","",list.files("results/figdata/p22500",pattern = "figdata.\\d+",full.names = FALSE))) %>% sort

for(pset in psets_to_process){

  filename <- file.path("results/figdata/p22500",paste("figdata.",sprintf(file_number_format,pset),".Rdata",sep=""))
  tryCatch({
    load(filename)

    h[[pset]] <- figdata[[3]]
    tf[[pset]] <- figdata[[1]]
    ar[[pset]] <- figdata[[2]]
    hd[[pset]] <- figdata[[4]]
    inps[[pset]] <- figdata[[5]]
    arp[[pset]] <- figdata[[6]]
    hivinc[[pset]] <- figdata[[7]]
    rm(figdata)

  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

save(h,tf,ar,hd,u5m,inps,arp,file="results/figdata/p22500/figdata_all.Rdata")
```
### Range of Inputs

@@

save(x,best.fitting.model,full.model.formula,file="results/best_fitting_model.RData")
```

@@

```{r, ori-fig-4, fig.cap="Original figure 4", echo=FALSE}
knitr::include_graphics("figure4-ori.png")
```

```{r, ori-fig-5, fig.cap="Original figure 5", echo=FALSE}
knitr::include_graphics("figure5-ori.png")
```

The figure 4 (Fig. \@ref(fig:ori-fig-4)) and 5 (Fig. \@ref(fig:ori-fig-5)) in the original article were created by applying the best-performing model as well as an independent model (REF) to empirical data from 2010 DHS for Malawi and Tanzania respectively to estimate U5M of different age groups of population. By comparing differences in U5M estimates of unadjusted data and adjusted data by two models, the authors suggested that the new model was able to correct the bias from indirect estimates.

With the new simulation data and best-fitting model, we re-generated both figures by using R scripts provided (Fig. \@ref(fig:rep-fig-4) and  Fig. \@ref(fig:rep-fig-5)).

@@

```{r, indirect-estimates, include=FALSE}

@@

```{r, Ward-Zaba-model, include=FALSE}
wz_model <- function(dataset, fig_country, iep){

  # Ward & Zaba (2008) coefficients
  a=c(.1134,.1202,.1107,.1720,.1749,.1504,.2552)
  b=c(-.0226,-.0438,-.0882,-.1412,-.1758,-.1849,-.3269)
  c=c(-.1112,-.0978,-.0373,.0163,.042,.1807,.5594)

@@

```{r, best-performing-model, include=FALSE}
best_model <- function(iep, fig_name){

@@

```{r, figure-plot, include=FALSE}
plot_fig <- function(fig_name, fig_country){
  # Load empirical data
  fv <- fread("./data/facevalidity.csv") %>% set_colnames(c("country","agegroup", "ceb","cs","tfr2010","tfr2000","hiv1990", "hiv2000", "hiv2010","art2005","art2006","art2007","art2008","art2009","art2010","art2011","art_prev2005","art_prev2006","art_prev2007","art_prev2008","art_prev2009","prop15to19_2010","prev15to19_2010","hiv2005","hiv2006","hiv2007","hiv2008","hiv2009")) %>% mutate(cd=ceb-cs)

@@

  # estimate from Ward & Zaba (2008) model
  fiveq0WZ <- wz_model(fv, fig_country, iep)

@@

```{r, rep-fig-4, fig.cap="Replication of figure 4 with new simulation data", echo=FALSE， fig.align='center'}

@@

```{r, rep-fig-5, fig.cap="Replication of figure 5 with new simulation data", echo=FALSE， fig.align='center'}

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Methods

```{r, load-packages}

@@

For the calculation of bias, we used the formula provided in the original paper. 
$$Relative bias = \frac{IE_{survivors} - IE_{survivors,HIV deaths}}{IE_{survivors,HIV deaths}}$$

$$Absolute bias = IE_{survivors} - IE_{survivors,HIV deaths}$$
According to [@quattrochi2019], $IE_{survivor}$ was the estimates of U5M based on living women in 2010, and $IE_{survivors,HIV deaths}$ was the estimates of U5M based on living women in 2010 and women died from HIV/AIDS before 2010 given the condition that they would have been 15-49 in 2010 if they had survived. 
$a_{d}$

## Simulation Construction 
### Range of Inputs

@@

save(x,best.fitting.model,full.model.formula,file="results/best_fitting_model.RData")

@@

```{r, ori-fig-4, fig.cap="Original figure 4"}
knitr::include_graphics("figures/figure4-ori.png")
```

```{r, ori-fig-4, fig.cap="Original figure 4"}
knitr::include_graphics("figures/figure5-ori.png")
```

The figure 4 (\@ref(fig:ori-fig-4)) and 5 (\@ref(fig:ori-fig-5)) in the original article were created by applying the best-performing model as well as an independent model (REF) to empirical data from 2010 DHS for Malawi and Tanzania respectively to estimate U5M of different age groups of population. By comparing differences in U5M estimates of unadjusted data and adjusted data by two models, the authors suggested that the new model was able to correct the bias from indirect estimates.

With the new simulation data and best-fitting model, we re-generated both figures by using R scripts provided (\@ref(fig:rep-fig-4) and \@ref(fig:rep-fig-5)).

@@

```{r, indirect-estimates}

@@

```{r, Ward-Zaba-model}
wz_model <- function(iep){
  # Ward & Zaba (2008) coefficients
  a=c(.1134,.1202,.1107,.1720,.1749,.1504,.2552)
  b=c(-.0226,-.0438,-.0882,-.1412,-.1758,-.1849,-.3269)
  c=c(-.1112,-.0978,-.0373,.0163,.042,.1807,.5594)

@@

```{r, best-performing-model}
best_model <- function(iep, fig_name){

@@

```{r, figure-plot}
plot_fig <- function(fig_name, fig_country){
  # Load empirical data
  fv <- fread("./data/facevalidity.csv") %>% set_colnames(c("country","agegroup", "ceb","cs","tfr2010","tfr2000","hiv1990", "hiv2000", "hiv2010","art2005","art2006","art2007","art2008","art2009","art2010","art2011","art_prev2005","art_prev2006","art_prev2007","art_prev2008","art_prev2009","prop15to19_2010","prev15to19_2010","hiv2005","hiv2006","hiv2007","hiv2008","hiv2009")) %>% mutate(cd=ceb-cs)

@@

  # estimate from Ward & Zaba (2008) model
  fiveq0WZ <- wz_model(iep)

@@

```{r, rep-fig-4, fig.cap="Replication of figure 4 with new simulation data"}

@@

```{r, rep-fig-5, fig.cap="Replication of figure 5 with new simulation data"}
